---
title: "Getting Data from Databases"
author: "Tony Yao-Jen Kuo"
output:
  revealjs::revealjs_presentation:
    highlight: pygments
    reveal_options:
      slideNumber: true
      previewLinks: true
---

# Introduction to Database

## What is database?

- DBMS
- Databases(Documents)
- Tables(Objects)
- Rows and columns(Keys and values)

## Why are we using databases?

It is hard to **scale** if we use **files** as the storage format of for our data.

## Databases are scalable in four dimensions

- C(reate)
- R(ead)
- U(pdated)
- D(elete)

## What is DBMS?

Short for Database Management System.

- Commercial License
  - Microsoft SQL Server
  - Oracle
  - IBM DB2
- Non-commercial License
  - MySQL
  - PostgreSQL
  - MongoDB

## What is SQL?

- Pronounces as **se-quel**
- Short for **Structured Query Language**
- The language used to query data stored in databases

## Another dimension viewing DBMS

- SQL
  - Microsoft SQL Server
  - Oracle
  - IBM DB2
  - MySQL
  - PostgreSQL
- NoSQL
  - MongoDB
  - Firebase
  
## SQL DBMS stores data in **table** formats

|position|player|team|season|
|--------|------|----|------|
|PG|Ron Harper|Chicago Bulls|1995-96|
|SG|Michael Jordan|Chicago Bulls|1995-96|
|SF|Scottie Pippen|Chicago Bulls|1995-96|
|PF|Dennis Rodman|Chicago Bulls|1995-96|
|C|Luc Longley|Chicago Bulls|1995-96|

## Whilist NoSQL DBMS stores data in JSON formats

```
{
  'team': 'Chicago Bulls',
  'season': '1995-96',
  'players': [
    {'PG': 'Ron Harper'},
    {'SG': 'Michael Jordan'},
    {'SF': 'Scottie Pippen'},
    {'PF': 'Dennis Rodman'},
    {'C': 'Luc Longley'}
  ]
}
```

## Choosing a DBMS that fits your requirements

- SQL is more concrete, but less flexible
- NoSQL is more flexible, but less concrete

# Connecting to a database with Python and R

## What are we gonna do?

Connecting to databases with external applications:

- Python
- R

## Databases used in class

|Type|DBMS|Cloud Provider|
|----|----|--------------|
|SQL|MySQL|AWS|
|NoSQL|Firebase|GCP|

## Development environments for Python and R

- Python: [Google Colab](https://colab.research.google.com/)
- R: [R](https://cran.r-project.org/) and [RStudio](https://www.rstudio.com/products/rstudio/download/)

## There is great variety when it comes to connection

- We've got different external applications
- We've also got different cloud service providers
- So it is basically a case-by-case situation

## Dealing with problems

- Commercial license: ask for tech support directly
- Non-commercial: reading documentation or asking for help

## The four requirements for connecting to databases

- host
- port
- username
- password

## Using module/package to establish a connector

Find the right module/package for SQL/NoSQL and Python/R, respectively.

|Programming Language|DBMS|Module/Package|
|--------------------|----|--------------|
|Python|MySQL|sqlalchemy/pymysql/pandas|
|Python|Firebase|firebase_admin|
|R|MySQL|RMySQL|
|R|Firebase|fireData|

# MySQL: Python

## Four requirements

- host: rsqltrain.ced04jhfjfgi.ap-northeast-1.rds.amazonaws.com
- port: 3306
- username: trainstudent
- password: csietrain

## Installing Python modules

Installing required modules before connecting.

```{bash eval = FALSE}
pip install --upgrade sqlalchemy pymysql pandas
```

## Creating a table

This should be granted to an administrator.

## Scripts for creating a table

```{python eval = FALSE}
import pandas as pd
from sqlalchemy import create_engine

csv_url = "https://storage.googleapis.com/ds_data_import/chicago_bulls_1995_1996.csv"
chicago_bulls = pd.read_csv(csv_url)
host = "YOURHOST" # Your own AWS RDS Enpoint
port = 3306
dbname = "YOURDBNAME" # Your own database name
user = "YOURUSERNAME" # Your own username
password = "YOURPASSWORD" # Your own password

engine = create_engine('mysql+pymysql://{user}:{password}@{host}:{port}/{dbname}'.format(user=user, password=password, host=host, port=port, dbname=dbname))
chicago_bulls.to_sql('chicago_bulls', engine, index=False, if_exists='replace')
```

## Importing a whole table from MySQL

`pd.read_sql_table()`

```{python eval = FALSE}
import pandas as pd
from sqlalchemy import create_engine

host = "rsqltrain.ced04jhfjfgi.ap-northeast-1.rds.amazonaws.com"
port = 3306
dbname = "nba"
user = "trainstudent"
password = "csietrain"

engine = create_engine('mysql+pymysql://{user}:{password}@{host}:{port}/{dbname}'.format(user=user, password=password, host=host, port=port, dbname=dbname))
chicago_bulls = pd.read_sql_table('chicago_bulls', engine)
chicago_bulls
```

## Importing data via a standard SQL query

`pd.read_sql_query()`

```{python eval = FALSE}
import pandas as pd
from sqlalchemy import create_engine

host = "rsqltrain.ced04jhfjfgi.ap-northeast-1.rds.amazonaws.com"
port = 3306
dbname = "nba"
user = "trainstudent"
password = "csietrain"

engine = create_engine('mysql+pymysql://{user}:{password}@{host}:{port}/{dbname}'.format(user=user, password=password, host=host, port=port, dbname=dbname))
sql_statement = """
  SELECT * 
  FROM chicago_bulls 
  WHERE Player IN ('Michael Jordan', 'Scottie Pippen', 'Dennis Rodman');
"""
trio = pd.read_sql_query(sql_statement, engine)
trio
```

# MySQL: R

## Installing R package

```{r eval = FALSE}
install.packages("RMySQL")
```

## Creating a table

This should be granted to an administrator.

## Scripts for creating a table

```{r eval = FALSE}
library(DBI)

csv_url <- "https://storage.googleapis.com/ds_data_import/boston_celtics_2007_2008.csv"
boston_celtics <- read.csv(csv_url)
host <- "YOURHOST" # Your own AWS RDS Enpoint
port <- 3306
dbname <- "YOURDBNAME" # Your own database name
user <- "YOURUSERNAME" # Your own username
password <- "YOURPASSWORD" # Your own password

engine <- dbConnect(RMySQL::MySQL(),
                    host = host,
                    port = port,
                    dbname = dbname,
                    user = user,
                    password = password
                    )
dbWriteTable(engine, name = 'boston_celtics', value = boston_celtics, overwrite = TRUE)
```

## Importing a whole table from MySQL

`dbReadTable()`

```{r eval = FALSE}
library(DBI)

host <- "rsqltrain.ced04jhfjfgi.ap-northeast-1.rds.amazonaws.com"
port <- 3306
dbname <- "nba"
user <- "trainstudent"
password <- "csietrain"

engine <- dbConnect(RMySQL::MySQL(),
                    host = host,
                    port = port,
                    dbname = dbname,
                    user = user,
                    password = password
                    )
boston_celtics <- dbReadTable(engine, name = 'boston_celtics')
View(boston_celtics)
```

## Importing data via a standard SQL query

`dbGetQuery()`

```{r eval = FALSE}
library(DBI)

host <- "rsqltrain.ced04jhfjfgi.ap-northeast-1.rds.amazonaws.com"
port <- 3306
dbname <- "nba"
user <- "trainstudent"
password <- "csietrain"

engine <- dbConnect(RMySQL::MySQL(),
                    host = host,
                    port = port,
                    dbname = dbname,
                    user = user,
                    password = password
                    )
sql_statement <- "SELECT * FROM boston_celtics WHERE Player IN ('Paul Pierce', 'Kevin Garnett', 'Ray Allen');"
gap <- dbGetQuery(engine, statement = sql_statement)
View(gap)
```

# Firebase: Python

## Installing Python module

```{bash eval = FALSE}
pip install firebase_admin
```

## Creating an object from Python

```{python eval = FALSE}
import firebase_admin
from firebase_admin import credentials
from firebase_admin import db
from requests import get

cred = credentials.Certificate('PATHTOYOURSERVICEACCOUNT') # Your own Service Account path
firebase_admin.initialize_app(cred, {
    'databaseURL' : 'YOURDATABASEURL' # Your own Firebase URL
})
# Creating object
json_url = 'https://storage.googleapis.com/ds_data_import/chicago_bulls_1995_1996.json'
chicago_bulls_dict = get(json_url).json()
root = db.reference()
root.child('chicago_bulls').push(chicago_bulls_dict)
```

## Importing object from Firebase

```{python eval = FALSE}
from firebase_admin import db

ref = db.reference('chicago_bulls')
chicago_bulls = ref.get()
chicago_bulls
```

# Firebase: R

## Installing R package

```{r eval=FALSE}
pkgs <- c("devtools", "jsonlite")
install.packages(pkgs)
devtools::install_github("Kohze/fireData")
```

## Creating an object from R

```{r eval=FALSE}
library(fireData)
library(jsonlite)

json_url <- "https://storage.googleapis.com/ds_data_import/boston_celtics_2007_2008.json"
boston_celtics_list <- fromJSON(json_url)
projectURL <- "YOURPROJECTURL"
upload(boston_celtics_list, projectURL = projectURL, directory = "boston_celtics")
```

## Importing object from Firebase

```{r eval=FALSE}
library(fireData)

projectURL <- "YOURPROJECTURL"
fileName <- "YOURDOCUMENTID"
boston_celtics_list <- download(projectURL = projectURL, fileName = fileName)
boston_celtics_list
```

# Summary

## In a nutshell

- How to authorize user
- Which module/package to use
- Handling data structures well

## Handling table and JSON

|Programming Language|Source|Data Structure|
|--------------------|------|--------------|
|Python|Table|`pd.DataFrame`|
|Python|JSON|`dict`|
|R|Table|`data.frame`|
|R|JSON|`list`|

# Reference

## Further readings

- <https://www.datainpoint.com/data-science-in-action/03-querying-databases.html>
- [Databases using R](https://db.rstudio.com/)
- <https://firebase.google.com/docs/?hl=zh-Tw>