---
title: "Manipulating Data"
author: "Tony Yao-Jen Kuo"
output:
  revealjs::revealjs_presentation:
    highlight: pygments
    reveal_options:
      slideNumber: true
      previewLinks: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Common data types

## There are 5 essential data types

|Python|R|
|------|-|
|`float`|`numeric` or `double`|
|`str`|`character`|
|`bool`|`logical`|
|`time.struct_time`|`POSIXct`|
|`None`|`NA` or `NULL`|

## Using functions to identify data types

- Using `type()` in Python
- Using `class()` or `typeof()` in R

## A quick review on basic data types

## Datetime in Python

```{python eval=FALSE}
from time import localtime, strftime

strftime("%Y-%m-%d %H:%M:%S", localtime())
```

## Datetime in R

```{r eval=FALSE}
as.character(Sys.time())
```

## References for `strptime`

- <https://docs.python.org/3/library/time.html#time.strftime>
- <https://stat.ethz.ch/R-manual/R-devel/library/base/html/strptime.html>

## Missing value in Python

```{python eval=FALSE}
print(type(None))
```

## Missing value in R

```{r eval=FALSE}
class(NA)
class(NULL)
```

## Difference between `NA` and `NULL` in R

```{r eval=FALSE}
my_vector <- c(1, 2, 3, NA)
sum(my_vector)
my_vector <- c(1, 2, 3, NULL)
sum(my_vector)
```

# Common data structures

## There are 4 essential data structures

|Python|R|
|------|-|
|`list`|`list`|
|`dict`|named `list`|
|`ndarray`|`vector`, `matrix`, `array`|
|`DataFrame`|`data.frame`|

## List in Python

```{python eval=FALSE}
my_list = [8.7, True, "Hello Python!"]
for e in my_list:
  print(type(e))
```

## List in R

```{r eval=FALSE}
my_list <- list(
  8.7,
  TRUE,
  "Hello R!"
)
for (e in my_list) {
  print(class(e))
}
```

## Dict in Python

```{python eval=FALSE}
my_dict = {
  "my_float": 8.7,
  "my_bool": True,
  "my_str": "Hello Python!"
}
```

## Named list in R

```{r eval=FALSE}
my_named_list <- list(
  my_float = 8.7,
  my_bool = TRUE,
  my_str = "Hello Python!"
)
```

## `ndarrays` in Python

```{python eval=FALSE}
import numpy as np

my_arr = np.array(range(11, 20))
print(type(my_arr))
```

## `vectors` in R

```{r eval=FALSE}
my_arr <- 11:19
class(my_arr)
```

## `DataFrame` in Python

```{python eval=FALSE}
import pandas as pd

numbers = [9, 23, 33, 91, 13]
players = ["Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley"]
df = pd.DataFrame()
df["number"] = numbers
df["player"] = players
df
```

## `data.frame` in R

```{r}
numbers <- c(9, 23, 33, 91, 13)
players <- c("Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley")
df <- data.frame(number = numbers, player = players, stringsAsFactors = FALSE)
View(df)
```

# Manipulating Dataframes

## There are six basic manipulation skills

- `filter`
- `select`
- `arrange`
- `mutate`
- `summarise`
- `group by`

## Dataframes we might use

- 1995-96 Chicago Bulls roster
- Lightweight `gapminder`

## The story behind `gapminder` data

<https://youtu.be/jbkSRLYSojo>

## Filtering dataframes in Python

```{python}
import pandas as pd

numbers = [9, 23, 33, 91, 13]
players = ["Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley"]
df = pd.DataFrame()
df["number"] = numbers
df["player"] = players
df.index = ["PG", "SG", "SF", "PF", "C"]
is_trio = df["number"].isin([23, 33, 91])
print(is_trio)
df[is_trio]
```

## Filtering dataframes in R

```{r}
library(dplyr)

numbers <- c(9, 23, 33, 91, 13)
players <- c("Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley")
df <- data.frame(number = numbers, player = players, stringsAsFactors = FALSE)
df %>% 
  filter(number %in% c(23, 33, 91))
```

## Selecting from dataframes in Python

```{python}
import pandas as pd

numbers = [9, 23, 33, 91, 13]
players = ["Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley"]
df = pd.DataFrame()
df["number"] = numbers
df["player"] = players
print(df["player"])
print(type(df["player"]))
df[["player", "number"]]
```

## Selecting from dataframes in R

```{r}
library(dplyr)

numbers <- c(9, 23, 33, 91, 13)
players <- c("Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley")
df <- data.frame(number = numbers, player = players, stringsAsFactors = FALSE)
df %>% 
  select(player, number)
```

## Arranging dataframes in Python

```{python}
import pandas as pd

numbers = [9, 23, 33, 91, 13]
players = ["Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley"]
df = pd.DataFrame()
df["number"] = numbers
df["player"] = players
print(df.sort_index(ascending=False))
print(df.sort_values(number))
```

## Arranging dataframes in R

```{r}
library(dplyr)

numbers <- c(9, 23, 33, 91, 13)
players <- c("Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley")
df <- data.frame(number = numbers, player = players, stringsAsFactors = FALSE)
df %>% 
  arrange(number)
```

## Mutating dataframes in Python

```{python}
import pandas as pd

numbers = [9, 23, 33, 91, 13]
players = ["Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley"]
df = pd.DataFrame()
df["number"] = numbers
df["player"] = players
df["last_name"] = df["player"].map(lambda x: x.split()[1])
df
```

## Mutating dataframes in R

```{r}
players <- c("Ron Harper", "Michael Jordan", "Scottie Pippen", "Dennis Rodman", "Luc Longley")
df <- data.frame(number = numbers, player = players, stringsAsFactors = FALSE)
get_last_name <- function(x) {
  split_lst <- strsplit(x, split = " ")
  name_length <- length(split_lst[[1]])
  last_name <- split_lst[[1]][name_length]
  return(last_name)
}
df$last_name <- sapply(df$player, FUN = get_last_name)
View(df)
```

## Summarising dataframes in Python

```{python}
import pandas as pd

csv_url = "https://storage.googleapis.com/learn_pd_like_tidyverse/gapminder.csv"
df = pd.read_csv(csv_url)
df[df.year == 2007]["pop"].sum()
```

## Summarising dataframes in R

```{r}
library(dplyr)

csv_url <- "https://storage.googleapis.com/learn_pd_like_tidyverse/gapminder.csv"
df <- read.csv(csv_url, stringsAsFactors = FALSE)
df %>% 
  filter(year == 2007) %>% 
  summarise(ttl_pop = sum(as.numeric(pop)))
```

## Grouping values of dataframe in Python

```{python}
import pandas as pd

csv_url = "https://storage.googleapis.com/learn_pd_like_tidyverse/gapminder.csv"
df = pd.read_csv(csv_url)
grouped = df[df.year == 2007].groupby("continent")
grouped["pop"].sum()
```

## Grouping value of dataframe in R

```{r}
library(dplyr)

csv_url <- "https://storage.googleapis.com/learn_pd_like_tidyverse/gapminder.csv"
df <- read.csv(csv_url, stringsAsFactors = FALSE)
df %>% 
  filter(year == 2007) %>% 
  group_by(continent) %>% 
  summarise(ttl_pop = sum(as.numeric(pop)))
```

# Summary

## In a nutshell

- Essential data types
- Essential data structures
  - Array-like
  - Hash-like
  - Table-like
- Manipulating table-like dataframes

# Reference

## Further readings

- <https://www.datainpoint.com/data-science-in-action/06-handling-data-structures.html>
- <https://www.datainpoint.com/data-science-in-action/07-manipulating-data-frames-basically.html>
- <https://www.datainpoint.com/data-science-in-action/08-manipulating-data-frames-advancingly.html>